The Creation of a Programming Language

### 2022-10-29 paper.tex
spellchecking

% Specify the version
\newcommand{\ver}[1]{
    \expandafter\ifstrequal\expandafter{\jobname}{paper}
    {#1}{\begin{center}-Removed, see full version-\end{center}}
}

@@

    % Include Mandelbrot on title page
    \begin{figure}[H]
        \centering
        \paste{paper/mandelbrot.tex}
        \caption{ASCII representation of the Mandelbrot Set, generated with \name{}}
        \label{f_mandelbrot}
    \end{figure}

@@

\section{Introduction}
The importance of computers in today's world is unquestioned.
But computers are not just the metals and semiconductors they are built from,
but would be unusable without any software running on them.
Creating programs, no matter if games, web browsers or even operating systems,
involve the use of programming languages.
Those serve as a bridge between the human mind and its imagination, and
the capabilities of computer hardware and are a huge time-saver when compared
to writing machine-native code, i.e. not using a programming language.
By making use of just the term \emph{language} this paper will refer to programming
languages specifically. Sources are mentioned fater cohesive paragraphs in
square brackets, an index can be found in the \emph{Bibliography} on page
\pageref{bibliography}.

Since the year 1973, when the first programming language was invented, the
research in programming languages has fastly advanced. But the goal has stayed
the same, to make programming as intuitive and easy, yet fast and efficient as
possible.

@@

\section{Implementation}
\subsection{Used Software}
At the heart of making any project in the C programming language
stands the C compiler. This project uses
\emph{gcc}, the GNU Compiler Collection, which, as the name suggests, also 
supports more languages than solely C. Substantiating the development process is 
\emph{GNU Make}, a general-purpose build system. It handles tasks like building
the final executable program step-by-step from the project source code and directory clean-up.
The C ecosystem is very easy to use on and fundamentally supported by 
\emph{GNU+Linux}, the operating system used to program on.
The code editor used was \emph{Microsoft Visual Studio Code}, providing great
help with built-in git integration and auto-completion features.
The tools \emph{Valgrind} and \emph{gdb}, the GNU Project Debugger were of great help
regarding the finding of errors and memory leaks, see section \ref{memleaks} for details.
\emph{Git} is the project management and version tracking software used to back up and share
the project files on \emph{GitHub.com}, a Microsoft hosted git repository and
web front-end to store and collaborate on source code.

@@

The problem to solve in this paper is concisely to make a computer understand
human-written and -readable text, called source code, and follow the instructions
it proposes.
Most interpreters follow a plain scheme to go about this problem, but the
actual implementation of each of the steps can be heavily customized.
Here is the problem, broken down into three steps as commonly done in other
popular programming languages:
\begin{enumerate}\pagelabel{simple_interpreter}
    \item To create meaningful groups of characters from the input string. This
        process is known as tokenization and is done with the component called
        tokenizer or lexer. As an example, the string \emph{ab + cd} could result
        in three tokens, namely \emph{ab}, \emph{+} and \emph{cd}. In this case, space characters
        are simply ignored and could be left out, unlike in the English language where spaces are relevant. 
        This is only an example and the grouping of characters could be done 
        in any thinkable, logical way.
    \item To order the tokens into parse-tree, also called syntax-tree, as
        specified by rules. These rules are called a grammar.
        Creating an order is important, as some calculations depend on others, e.g. multiplication
        is always done before addition. An IT data structure that can accommodate
        such requirements is the binary tree. A binary tree consists of tree-nodes
        that each have a maximum of two other tree nodes as their children, and
        a tree node whose child they are, the parent node.
    \item To interpret the parse tree, traversing through it and following the
        instructions each of the tree-nodes represent. The outcome
        will be the result of the computation the program describes.
\end{enumerate}

@@

    \item \emph{Functions.} In programming, functions or methods are
        a set of code that accomplish a certain task. They take in data,
        process it and return a result. Calling a function means to evaluate it for a given input.
        In C, every function has a return type, a
        function name, arguments and a function body in that order.
        Here is an example of a C function:
        \code{pseudocode/function.c}{1}{999}
        This would be an example of a main function, similar to the one
        in \hrc{src/main.c}. It returns a value
        of type 'int', has two arguments, 'argc' and 'argv', and a function body
        which, in the case of the example above, prints the value of the 'argc' argument.

@@

Following is a very quick summary of the few 
things happening in the \hrc{src/main.c} file.
First, the \hrc{src/interpreter/interpreter.h} header file is included. 
The main function then decides whether to
read a program from a file or from standard input. 
At last, the interpret function
from the said included header is called with the selected input.

@@

The \emph{interpret} function calls the parser to generate a syntax-tree from tokens retrieved
from the tokenizer implemented in \hrc{src/interpreter/parser.c} and 
\hrc{src/interpreter/tokenizer.c} respectively.

After that, the names of variables are
mapped to addresses in memory where their value is stored. Doing this before
bytecode compilation has a performance benefit. 
The mapping is acheived with a custom hashtable datastructure with average time
complexity of O(log n), n being the total amount of variables a \name{} program is using.
The hashtable implementation used in \hrc{src/interpreter/memory/hashtable.c}                   % TODO: maybe its own paragraph! interpolation search vs binary search
uses binary search to locate elements variable-sized arrays in the hashtable's buckets.
This is because for such few variables that are used in a standard \name{} program,
an interpolation search algorithm is actually slower.
For collision detection, i.e. to circumvent problems arising when two variable names have the
same hash, a technique called separate chaining is used. In this approach, linked lists with
key-value pairs are used for each array index of the buckets and variable names are compared directly
to find exact matches.
[\cte{separate_chaining}]

@@

Here is an example of tokenization:
An input of \emph{a.b==23.4} would result in the following tokens according
to the tokenizer implementation of this paper:
\begin{enumerate}
    \item Token of type FIELD with content \emph{a.b}.
    \item Token of type SYMBOL with content \emph{==}.
    \item Token of type NUMBER with content \emph{23.4}.
\end{enumerate}
FIELD type tokens are later replaced by REFERENCE tokens, effectively mapping
variable names to addresses in memory.
SYMBOL type tokens are mapped to operators such as plus or minux by \hrc{src/implementation/mappings/mapop.c}
during parsing. NUMBER type tokens are replaced by an actual number type, which
is C's 'long double' type as specified in \hrc{src/interpreter/bytecode.h}.

While tokenizers can be programmed using regular expressions, 
this paper uses a more simple finite-state machine tokenizer.
It reads character by character, determining the token type as it goes.
It stops when there is a token thad does no longer match the current token type.
Tokens are each requested and integrated into a syntax-tree by the parser as described in the next
section.

@@

The difficult part of implementing a parser is precedence. Precedence together
with associativity define the order of execution of statements. A simple example
would be to perform multiplication before addition, e.g. \emph{2+3*4} would
result in \emph{14} and is not the same as \emph{(2+3)*4}.
Assosiativity is either 'Left to Right' or 'Right to Left'. An example of associativity is
\emph{20/10/2}. Both in C as well as in \name{}, the division operator is
left-to-right associative, meaning the stated expression is the same as
\emph{(20/10)/2} and not as \emph{20/(10/2)}. Both precedence 
and associativity are properties that are defined in the file \hrc{mapop.gpref} for
each operator.
In the end the parse tree contains all values and 
operations as well as information about the order of execution.

The parser implementation in \hrc{src/interpreter/parser.c} may be difficult to
read and understand, which is why the most important functions are explained here:
\begin{itemize}
    \item \emph{advance.} This function reads the next token from the tokenizer.
    \item \emph{secondary.} This function reads tokens that will become the leaf nodes
        of the parse tree.
        It deals with simple values that make up the start of an expression, such as numbers,
        fields, strings and it handles brackets. 
        In addition, it can handle prefix operators, such as the unary plus or unary minus.
    \item \emph{expr.} The \emph{expr} function calls \emph{secondary} whose result acts
        as the basis for a new expression. It then repeatedly checks
        if an operator follows the secondary token and creates the syntax-tree
        with help of recursive mechanisms. The nature of this function also 
        removees the need for semi-colons,
        which are essential in languages like Java or C.
    \item \emph{parse.} The parse function is a small wrapper which only
        calls the \emph{expr} function a first time, initiating the parsing process.
\end{itemize}

Following is a piece of pseudocode to better illustrate the inner workings of the
parser which is based on a Pratt-parser. 
This is only an example and not completely accurate to the actual parser used
by this project.
% TODO: validate code
\code{pseudocode/parser.c}{1}{999}

@@

\subsubsection{The Runtime}
This is the last stage of interpreting program code as implemented in \name{}.
The runtime steps through the intermediary bytecode created by the bytecode
compiler and evaluates expressions by plugging in the provided values.
All available operators are implemented in 
\hrc{src/interpreter/processing/implementations.c}.

The evaluation of expressions works in a recursive manner. Here is an 
example of \name{}'s bytecode notation and how it is evaluated at runtime:
\begin{itemize}
    \item Assume the input in \name{}'s bytecode to resemble the following expression: 
        \expr{\emph{- 6 * 3 4}}
        In common infix notation, this would be equivalent to 
        \expr{\emph{(4 * 3) - 6}.}
        The question arises, how can an algorithm compute the result of this
        calculation? The answer is recursion.
    \item The runtime will read through the bytecode from left to right. First it will
        notice the subtraction operator and accordingly assume that two
        operands must follow.
    \item The function recursively calls itself in order to evaluate the
        right operand and notice the 6, a value and a leaf node in the parse tree. 
        The function simply returns the 6 to the subtraction expression.
    \item Again, the function calls itself to find the next operand.
        As it now reaches a multiplication operator,
        it recursively reads through that sub-expression's right and then left
        operator, which are 3 and 4 respectively.
        The result is computed and returned.
    \item Now the subtraction knows its right- and left-hand operators, 6 and 12.
        The subtraction is performed and the result, 6, is returned.
\end{itemize}

That is it. With all the operators and standard functions that have been implemented, entire programs
can be interpreted. Examples of such programs can be found in the \emph{examples} directory
of the source code.

@@

% TODO: add examples in /examples/
\name{} is built upon the implementation of a recursive Pratt
parser and a basic tokenizer. It uses a hash table to map variable names to their
memory addresses. There are no keywords, only a set of non-alphanumeric characters
representing all possible operations. Extending upon that, there are some
standard functions that enlargen the capabilities of the language.

@@

Following is a piece of example \name{} code, printing out the first ten numbers
of the fibonacci sequence, as well as a disection of how that same code is interpreted.
\code{examples/fibonacci.nts}{1}{999}
\begin{itemize}
\item The source code begins with the definition two variables \emph{a} and \emph{b}, which
    will be utilized to keep track of the data used to calculate fibonacci numbers.
    They are each assigned a number, zero and one respectively.
    Since the variable names \emph{a} and \emph{b} are not yet
    known to the interpreter, they are automatically allocated in the current scope
    and assigned a default value. That default value is then overwritten with the
    assignment operator.
\item Another variable \emph{count} is declared and defined as the number ten on line 4.
    This variable will be used to limit the amount of numbers the program outputs.
\item Now the function \emph{std.repeat} is called. This function is a
    standard library function, meaning it is included in the \name{} interpreter and
    written in C. Another function is passed as argument to the function call.
    This function will be repeated.
\item On line 7 and 8, some calculations are done.
    Remember, the equals sign stands for an assignment, not an
    equation. First, \emph{a} is assigned the sum of \emph{a} and \emph{b}. Then 
    \emph{b} is set to be the new \emph{a} minus \emph{b}, 
    which is in turn the same as \emph{a} before the reassignment on line 7.
\item On lines 10 and 11, the standard library function \emph{std.print} is used
    to print out the value of a and a line break. The backslash in the text
    that is print out signifies a so-called escape-sequence. The \emph{n} stands
    for a newline, but other letters such as a \emph{t} for a tabulator are possible.
\item Lastly, line 13 contains another calculation in which the variable \emph{count}
    is decremented by one. Since this is the last instruction in the inner function,
    it is simoultaneously its return value, e.g. the value of count is returned
    to the \emph{std.repeat} function. The \emph{std.repeat} function will
    repeat the inner function until the returned value is zero or undefined.
    This means the inner function is executed exacly ten times, because
    the initial value of count has to be decremented by one ten times until
    it equals zero.
\end{itemize}

\subsection{Name Shadowing}
Here is another example, which in contrast to the example above, focuses on an
original security concept implemented in \name{}.
\code{examples/exec.nts}{1}{999}
The code begins by creating the variable \emph{var}, which holds a string and
\emph{f}, which calls the function \emph{std.exec} with var as argument.
Line number seven holds a comment, indicated by the two slashes at the start of the line.
Comments are ignored by the \name{} interpreter and their purpose is to contain notes
and explanations inside the source code itself. Another type of commends are block-comments
which span across multiple lines, such as on lines 9 to 12.
On line seven, the \emph{std.exec} function is called for the first time. It runs the
command it is given as argument in the user's shell, which is most commonly \emph{cmd} on
Microsoft Windows and \emph{bash} on most Linux distributions. In this example, \emph{std.exec}
launches the Firefox browser, but it is capable of all kinds of things that can otherwise be
done in the shell, such as delete files or shut down the computer.

@@

This security concept enforces the use of a top-down programming model. First, safe-to-use aliases
to dangerous functions have to be created. Then, access to all those exploitable functions is
revoked. This way, running unknown code can only access data and functions that are being provided.
This contrasts the models of languages like Python. In Python, each seperate source file can
make use of the \emph{import} keyword to include other files. Files imported this way do not
have access or knowledge about the variables that exist in the file that includes them.
In \name{}, each imported file would instead have access to all variables and functions that
are made available to it, instead of having to include or import them seperately.
This leads to another problem, but that is what the next section will be all about.

@@

The name \name{} was chosen, because it implised this programming language
to be a learning project as well as a scripting language, which employ
high-level abstractions of underlying operations and interpret one line
at a time. The file name extension \emph{.nts} is used in the examples. This
file extension is not required, but was chosen as it stands for \emph{Not TypeScript}, because
the language neither enforces strict typing, nor is it Microsoft's \emph{TypeScript}.

@@

\subsubsection{Numbers}
Another goal that turned out not to be practical to acheive was to use just as much
memory for each variable as needed and not more. This means every variable would
have to be assigned a datatype, containing information about the size and possible values.
Taking C as a reference, the datatype \emph{char} is exactly one byte in size, and
can therefore be in exactly 2^8 or 256 different states. But in C, information about
the datatype can be discarded at compile time, whereas in an interpreter like \name{} such
information needs to persist throughout the program's lifetime.

\subsubsection{Memory Leaks}\label{memleaks}
Memory leaks happen when a program reserves memory to store data, but
never frees up the space for use for other programs, even though the data
has long been forgotten about and will no longer be accessed.
In \name{}, memory leaks are a major problem. Even though barely noticable in small
programs it interprets, \name{} does not have proper setups in place to free up
all used memory that is not used any longer.

Manx rogramming languages, such as Java and Python,
make use of Garbage-Collection, a strategy to track unused allocated memory and
free it up automatically. While \name{} also tries to make use of this technique,
there are many flaws introduced while developing the interpreter that hinder 
Garbage-Collection from working correctly.

@@

A tool that is helpful to find and fix memory leaks is Valgrind. It runs
the executable program against an input and finds the origins of unfreed data.

@@

Most widely used programming languages not only consist
of a language specification and a compiler or interpreter, but of many more
tools that surround themn.
Such tools can include a package-manager, with which people around the world
can share code. Examples would be \emph{npm} or \emph{cargo}, 
the 'Node Package Manager' for JavaScript applications and and Rust's package 
manager respectively.
Another example of a helpful tool is a debugger, helping developers
find the cause of errors, so-called bugs, in their programs.
This includes the tool gdb, the GNU Project Debugger, that was used in this
project to find errors in the \name{} implementation.
[\cte{npm}, \cte{cargo}]

On the other hand, having to deal with countless package-managers for each
language can be annoying for developers. A solution is to build upon and
extend already existing languages and their tools with e.g. a new syntax.
This is exactly what Kotlin does. Kotlin can produce Java Virtual Machine
compatible bytecode, can be transpiled into
JavaScript, or ECMAScript 5.1 to be more specific,
and it can be compiled to machine-native programs via LLVM. 
[\cte{kotlin_compile}]

@@

\subsubsection{Foreign Function Interfaces}
A foreign function interface, often abbreviated as FFI,
is a mechanism by which one programming language can call functions written 
in another. This makes a language more attractive, as it can be used with
existing libraries from another language, instead of functions having to be
re-written. Enabling C function calls often allows a certain language to
directly interact with system libraries or the operating system itsself.

Foreign function interfaces are available in many modern languages.
Examples are Java, Python and Rust.
Java enables users to call C, C++ and even assembly code with the Java Native
Interface, JNI for short, whereas within Python C function can
be run with the standard ctypes library.

@@

\subsubsection{Reflection}
Reflection, also called reflective programming, describes
the ability of a program to examine and modify its own structure and behavior 
at runtime. An example would be self-modifying code. This can easily be
achieved in assembly language, which inherently does not differentiate between
data and instructions.
Another example is Java, providing methods and classes in the java.lang.reflect
package that enable developers to examine and change properties or functions
of objects.
The advantage of reflection over not having it is that it allows for non-static
programs and program flows, effectively making smaller executables and more
resource efficient programs, as those programs can adjust themselves to run
differently under different conditions at runtime.
The easiest and simplest way to go about implementing Reflection in \name{} is
certainly to allow the interpretation of strings at runtime. That way, strings
holding program instructions can be created and modified and later executed.
[\cte{reflection}]

@@

\newpage\subsection{Bibliography}
\renewcommand*{\bibfont}{\normalsize}
\pagelabel{bibliography}\printbibliography[heading=none]

@@

\section{Declaration of Autonomy}
Hereby I, Florian Donnelly, declare to have written this paper, also including
the entire source code of \name{} by myself using resources listed under the
\emph{Bibliography} section on page \pageref{bibliography}.
% TODO: check correctness of all file paths!! e.g. src/...

>>>>

% Specify the version
\newcommand{\ver}[1]{
    \expandafter\ifstrequal\expandafter{\jobname}{paper}
    {#1}{}
}

@@

    % Include Mandelbrot on title page
    \ver{
    \begin{figure}[H]
        \centering
        \paste{paper/mandelbrot.tex}
        \caption{ASCII representation of the Mandelbrot Set, generated with \name{}}
        \label{f_mandelbrot}
    \end{figure}
    }

@@

\section{Introduction}
The importance of computers in today's world is unquestioned.
But computers are not just the metals and semiconductors they are built from,
but would be unusable without any software running on them.
Creating programs, no matter if games, web browsers or even operating systems,
involve the use of programming languages.
Those serve as a bridge between the human mind and its imagination, and
the capabilities of computer hardware and are a huge time-saver when compared
to writing machine-native code, i.e. not using a programming language.
By making use of just the term \emph{language} this paper will refer to programming
languages specifically. Sources are mentioned after cohesive paragraphs in
square brackets, an index can be found in the \emph{Bibliography} on page
\pageref{bibliography}.

Since the year 1973, when the first programming language was invented, the
research in programming languages has vastly advanced. But the goal has stayed
the same, to make programming as intuitive and easy, yet fast and efficient as
possible.

@@

\section{Implementation}
\subsection{Used Software}
At the heart of making any project in the C programming language
stands the C compiler. This project uses
\emph{gcc}, the GNU Compiler Collection, which, as the name suggests, also 
supports more languages than solely C. Substantiating the development process is 
\emph{GNU Make}, a general-purpose build system. It handles tasks like building
the final executable program step-by-step from the project source code and directory clean-up.
The C ecosystem is very easy to use on and fundamentally supported by 
\emph{GNU+Linux}, the operating system used to program on.
The code editor used was \emph{Microsoft Visual Studio Code}, providing great
help with built-in git integration and auto-completion features.
The tools \emph{Valgrind} and \emph{gdb}, the GNU Project Debugger were of great help
when trying to find of errors and memory leaks, see section \ref{memleaks} for details.
\emph{Git} is the project management and version tracking software used to back up and share
the project files on \emph{GitHub.com}, a Microsoft hosted git repository and
web front-end to store and collaborate on source code.

@@

The problem to solve in this paper is concisely to make a computer understand
human written and -readable text, called source code, and follow the instructions
it proposes.
Most interpreters follow a plain scheme to go about this problem, but the
actual implementation of each of the steps can be heavily customized.
Here is the problem, broken down into three steps as commonly done in other
popular programming languages:
\begin{enumerate}\pagelabel{simple_interpreter}
    \item To create meaningful groups of characters from the input string. This
        process is known as tokenization and is done with the component called
        tokenizer or lexer. As an example, the string \emph{ab + cd} could result
        in three tokens, namely \emph{ab}, \emph{+} and \emph{cd}. In this case, space characters
        are simply ignored and could be left out, unlike in the English language where spaces are relevant. 
        This is only an example and the grouping of characters could be done 
        in any thinkable, logical way.
    \item To order the tokens into parse-tree, also called syntax-tree, as
        specified by rules. These rules are called a grammar.
        Creating an order is important, as some calculations depend on others, e.g. multiplication
        is always done before addition. An IT data structure that can accommodate
        such requirements is the binary tree. A binary tree consists of tree-nodes
        that each have a maximum of two other tree nodes as their children, and
        a tree node whose child they are, the parent node.
    \item To interpret the parse tree, traversing through it and following the
        instructions each of the tree-nodes represent. The outcome
        will be the result of the computation the program describes.
\end{enumerate}

@@

    \item \emph{Functions.} In programming, functions or methods are
        a set of code that accomplish a certain task. They take in data,
        process it and return a result. Calling a function means to evaluate it for a given input.
        In C, every function has a return type, a
        function name, arguments and a function body in that order.
        Here is an example of a C function:
        \code{pseudocode/function.c}{1}{999}
        This would be an example of a main function, similar to the one
        in \hrc{src/main.c}. It returns a value
        of type \emph{int}, has two arguments, \emph{argc} and \emph{argv}, and a function body
        which, in the case of the example above, prints out the value of the \emph{argc} argument.

@@

Following is a very quick summary of the few 
things happening in the \hrc{src/main.c} file.
First, the \hrc{src/interpreter/interpreter.h} header file is included. 
The main function then decides whether to
read a program from a file or from standard input. 
At last, the function \emph{interpret}
from the said included header is called with the selected input.

@@

The function \emph{interpret} calls the parser to generate a syntax-tree from tokens retrieved
from the tokenizer implemented in \hrc{src/interpreter/parser.c} and 
\hrc{src/interpreter/tokenizer.c} respectively.

After that, the names of variables are
mapped to addresses in memory where their value is stored. Doing this before
bytecode compilation has a performance benefit. 
The mapping is achieved with a custom hash table data structure with average time
complexity of O(log n), n being the total amount of variables a \name{} program is using.
The hash table implementation used in \hrc{src/interpreter/memory/hashtable.c}                   % TODO: maybe its own paragraph! interpolation search vs binary search
uses binary search to locate elements variable-sized arrays in the hashtable's buckets.
This is because for such few variables that are used in a standard \name{} program,
an interpolation search algorithm is actually slower.
For collision detection, i.e. to circumvent problems arising when two variable names have the
same hash, a technique called separate chaining is used. In this approach, linked lists with
key-value pairs are used for each array index of the buckets and variable names are compared directly
to find exact matches.
[\cte{separate_chaining}]

@@

Here is an example of tokenization:
An input of \emph{a.b==23.4} would result in the following tokens according
to the tokenizer implementation of this paper:
\begin{enumerate}
    \item A token of type FIELD with content \emph{a.b}.
    \item A second token of type SYMBOL with content \emph{==}.
    \item A third token of type NUMBER with content \emph{23.4}.
\end{enumerate}
FIELD type tokens are later replaced by REFERENCE tokens, effectively mapping
variable names to addresses in memory.
SYMBOL type tokens are mapped to operators such as plus or minux by \hrc{src/implementation/mappings/mapop.c}
during parsing. NUMBER type tokens are replaced by an actual number type, which
is C's 'long double' type as specified in \hrc{src/interpreter/bytecode.h}.

While tokenizers can be programmed using regular expressions, 
this paper uses a simpler finite-state machine tokenizer.
It reads character by character, determining the token type as it goes.
It stops when there is a token that does no longer match the current token type.
Tokens are each requested and integrated into a syntax-tree by the parser as described in the next
section.

@@

The difficult part of implementing a parser is precedence. Precedence together
with associativity define the order of execution of statements. A simple example
would be to perform multiplication before addition, e.g. \emph{2+3*4} would
result in \emph{14} and is not the same as \emph{(2+3)*4}.
Associativity is either 'Left to Right' or 'Right to Left'. An example of associativity is
\emph{20/10/2}. Both in C and in \name{}, the division operator is
left-to-right associative, meaning the stated expression is the same as
\emph{(20/10)/2} and not as \emph{20/(10/2)}. Both precedence 
and associativity are properties that are defined in the file \hrc{mapop.gpref} for
each operator.
In the end the parse tree contains all values and 
operations as well as information about the order of execution.

The parser implementation in \hrc{src/interpreter/parser.c} may be difficult to
read and understand, which is why the most important functions are explained here:
\begin{itemize}
    \item \emph{advance.} This function reads the next token from the tokenizer.
    \item \emph{secondary.} This function reads tokens that will become the leaf nodes
        of the parse tree.
        It deals with simple values that make up the start of an expression, such as numbers,
        fields, strings and it handles brackets. 
        In addition, it can handle prefix operators, such as the unary plus or unary minus.
    \item \emph{expr.} The \emph{expr} function calls \emph{secondary} whose result acts
        as the basis for a new expression. It then repeatedly checks
        if an operator follows the secondary token and creates the syntax-tree
        with help of recursive mechanisms. The nature of this function also 
        removes the need for semicolons,
        which are essential in languages like Java or C.
    \item \emph{parse.} The parse function is a small wrapper which only
        calls the \emph{expr} function a first time, initiating the parsing process.
\end{itemize}

Following is a piece of pseudo code to better illustrate the inner workings of the
parser which is based on a Pratt-parser. 
This is only an example and not completely accurate to the actual parser used
by this project.
% TODO: validate code
\code{pseudocode/parser.c}{1}{999}

@@

\subsubsection{Program Evaluation}
This is the last stage of interpreting program code as implemented in \name{}.
The interpreter steps through the intermediary bytecode created by the bytecode
compiler and evaluates expressions by plugging in the provided values.
All available operators are implemented in 
\hrc{src/interpreter/processing/implementations.c}.

The evaluation of expressions works in a recursive manner. Here is an 
example of \name{}'s bytecode notation and how it is evaluated:
\begin{itemize}
    \item Assume the input in \name{}'s bytecode to resemble the following expression: 
        \expr{\emph{- 6 * 3 4}}
        In common infix notation, this would be equivalent to 
        \expr{\emph{(4 * 3) - 6}.}
        The question arises, how can an algorithm compute the result of this
        calculation? The answer is recursion.
    \item The interpreter will read through the bytecode from left to right. First it will
        notice the subtraction operator and accordingly assume that two
        operands must follow.
    \item The function recursively calls itself in order to evaluate the
        right operand and notice the 6, a value and a leaf node in the parse tree. 
        The function simply returns the 6 to the subtraction expression.
    \item Again, the function calls itself to find the next operand.
        As it now reaches a multiplication operator,
        it recursively reads through that sub-expression's right and then left
        operator, which are 3 and 4 respectively.
        The result is computed and returned.
    \item Now the subtraction knows its right- and left-hand operators, 6 and 12.
        The subtraction is performed and the result, 6, is returned.
\end{itemize}

That is it. With all the operators and standard functions that have been implemented, entire programs
can be interpreted. Examples of such programs can be found in the directory
\emph{examples} of the source code.

@@

% TODO: add examples in /examples/
\name{} is built upon the implementation of a recursive Pratt
parser and a basic tokenizer. It uses a hash table to map variable names to their
memory addresses. There are no keywords, only a set of non-alphanumeric characters
representing all possible operations. Extending upon that, there are some
standard functions that enlarge the capabilities of the language.

@@

Following is a piece of example \name{} code, printing out the first ten numbers
of the Fibonacci sequence, as well as a dissection of how that same code is interpreted.
\code{examples/fibonacci.nts}{1}{999}
\begin{itemize}
\item The source code begins with the definition two variables \emph{a} and \emph{b}, which
    will be utilized to keep track of the data used to calculate Fibonacci numbers.
    They are each assigned a number, zero and one respectively.
    Since the variable names \emph{a} and \emph{b} are not yet
    known to the interpreter, they are automatically allocated in the current scope
    and assigned a default value. That default value is then overwritten with the
    assignment operator.
\item Another variable \emph{count} is declared and defined as the number ten on line 4.
    This variable will be used to limit the amount of numbers the program outputs.
\item Now the function \emph{std.repeat} is called. This function is a
    standard library function, meaning it is included in the \name{} interpreter and
    written in C. Another function is passed as argument to the function call.
    This function will be repeated.
\item On line 7 and 8, some calculations are done.
    Remember, the equals sign stands for an assignment, not an
    equation. First, \emph{a} is assigned the sum of \emph{a} and \emph{b}. Then 
    \emph{b} is set to be the new \emph{a} minus \emph{b}, 
    which is in turn the same as \emph{a} before the reassignment on line 7.
\item On lines 10 and 11, the standard library function \emph{std.print} is used
    to print out the value of a and a line break. The backslash in the text
    that is print out signifies a so-called escape sequence and will be replaced with
    piece of text that can not be entered directly. The \emph{n} stands
    for a line break, but other letters such as a \emph{t} for a tabulator are possible.
\item Lastly, line 13 contains another calculation in which the variable \emph{count}
    is decremented by one. Since this is the last instruction in the inner function,
    it is simultaneously its return value, e.g. the value of count is returned
    to the \emph{std.repeat} function. The \emph{std.repeat} function will
    repeat the inner function until the returned value is zero or undefined.
    This means the inner function is executed exactly ten times, because
    the initial value of count has to be decremented by one ten times until
    it equals zero.
\end{itemize}

\subsection{Name Shadowing}
Here is another example, which in contrast to the example above, focuses on an
original security concept implemented in \name{}.
\code{examples/exec.nts}{1}{999}
The code begins by creating the variable \emph{var}, which holds a string and
\emph{f}, which calls the function \emph{std.exec} with var as argument.
Line number seven holds a comment, indicated by the two slashes at the start of the line.
Comments are ignored by the \name{} interpreter and their purpose is to contain notes
and explanations inside the source code itself. Another type of commends are block-comments
which span across multiple lines, such as on lines 9 to 12.
On line seven, the \emph{std.exec} function is called for the first time. It runs the
command it is given as argument in the user's shell, which is most commonly \emph{cmd.exe} on
Microsoft Windows and \emph{bash} on most Linux distributions. In this example, \emph{std.exec}
launches the Firefox browser, but it is capable of all kinds of things that can otherwise be
done in the shell, such as delete files or shut down the computer.

@@

This security concept enforces the use of a top-down programming model. First, safe to use aliases
to dangerous functions have to be created. Then, access to all those exploitable functions is
revoked. This way, running unknown code can only access data and functions that are being provided.
This contrasts the models of languages like Python. In Python, each separate source file can
make use of the \emph{import} keyword to include other files. Files imported this way do not
have access or knowledge about the variables that exist in the file that includes them.
In \name{}, each imported file would instead have access to all variables and functions that
are made available to it, instead of having to include or import them separately.
This leads to another problem, but that is what the next section will be all about.

@@

The name \name{} was chosen, because it implied this programming language
to be a learning project as well as a scripting language, which employ
high-level abstractions of underlying operations and interpret one line
at a time. The file name extension \emph{.nts} is used in the examples. This
file extension is not required, but was chosen as it stands for \emph{Not TypeScript}, because
the language neither enforces strict typing, nor is it Microsoft's \emph{TypeScript}.

@@

\subsubsection{Numbers}
Another goal that turned out not to be practical to achieve was to use just as much
memory for each variable as needed and not more. This means every variable would
have to be assigned a datatype, containing information about the size and possible values.
Taking C as a reference, the datatype \emph{char} is exactly one byte in size, and
can therefore be in exactly 2\^8 or 256 different states. But in C, information about
the datatype can be discarded at compile time, whereas in an interpreter like \name{} such
information needs to persist throughout the program's lifetime.

\subsubsection{Memory Leaks}\label{memleaks}
Memory leaks happen when a program reserves memory to store data, but
never frees up the space for use for other programs, even though the data
has long been forgotten about and will no longer be accessed.
In \name{}, memory leaks are a major problem. Even though barely noticeable in small
programs it interprets, \name{} does not have proper setups in place to free up
all used memory that is not used any longer.

Many programming languages, such as Java and Python,
make use of Garbage-Collection, a strategy to track unused allocated memory and
free it up automatically. While \name{} also tries to make use of this technique,
there are many flaws introduced while developing the interpreter that hinder 
Garbage-Collection from working correctly.

@@

A tool that is helpful to find and fix memory leaks is \emph{Valgrind}. It runs
the executable program against an input and finds the origins of not freed data.

@@

Most widely used programming languages not only consist
of a language specification and a compiler or interpreter, but of many more
tools that surround them.
Such tools can include a package-manager, with which people around the world
can share code. Examples would be \emph{npm} or \emph{cargo}, 
the 'Node Package Manager' for JavaScript applications and and Rust's package 
manager respectively.
Another example of a helpful tool is a debugger, helping developers
find the cause of errors, so-called bugs, in their programs.
This includes the tool gdb, the GNU Project Debugger, that was used in this
project to find errors in the \name{} implementation.
[\cte{npm}, \cte{cargo}]

On the other hand, having to deal with countless package-managers for each
language can be annoying for developers. A solution is to build upon and
extend already existing languages and their tools with e.g. a new syntax.
This is exactly what Kotlin does. Kotlin can produce Java Virtual Machine
compatible bytecode, can be transcompiled, e.g. translated, into
JavaScript, or ECMAScript 5.1 to be more specific,
and it can be compiled to machine-native programs via LLVM. 
[\cte{kotlin_compile}]

@@

\subsubsection{Foreign Function Interfaces}
A foreign function interface, often abbreviated as FFI,
is a mechanism by which one programming language can call functions written 
in another. This makes a language more attractive, as it can be used with
existing libraries from another language, instead of functions having to be
re-written. Enabling C function calls often allows a certain language to
directly interact with system libraries or the operating system itself.

Foreign function interfaces are available in many modern languages.
Examples are Java, Python and Rust.
Java enables users to call C, C++ and even assembly code with the Java Native
Interface, JNI for short, whereas within Python C function can
be run with the standard library \emph{ctypes}.

@@

\subsubsection{Reflection}
Reflection, also called reflective programming, describes
the ability of a program to examine and modify its own structure and behavior 
at runtime. An example would be self-modifying code. This can easily be
achieved in assembly language, which inherently does not differentiate between
data and instructions.
Another example is Java, providing methods and classes in the java.lang.reflect
package that enable developers to examine and change properties or functions
of objects.
The easiest and simplest way to go about implementing Reflection in \name{} is
certainly to allow the interpretation of strings at runtime. That way, strings
holding program instructions can be created and modified and later executed.
[\cte{reflection}]

@@

\subsection{Bibliography}
\renewcommand*{\bibfont}{\normalsize}
\pagelabel{bibliography}\printbibliography[heading=none]

@@

\section{Declaration of Authenticity}
Hereby I, Florian Donnelly, declare to have written this paper independently, 
also including the entire source code of \name{} by myself using resources 
listed under the \emph{Bibliography} section on page \pageref{bibliography}.
% TODO: check correctness of all file paths!! e.g. src/...
